Lossy compression of DNA sequencing quality data

### 2015-04-21 manuscript/q-compression.tex
Text clean up.

\section{Results:}
Lossy compression of quality information can greatly decrease storage
and memory requirements with little discernible effects on subsequent
analysis results. The four compression strategies in this study were
able to produce similar results to those obtained with uncompressed
quality sequences in terms of quality control, genome assembly, and
alignment of short read to a reference sequence.

@@

If quality values are modeled compression is conceivably possible by
replacing the original quality values by a representation of the
model. For example, quality values can be conceptualized as bivariate
data with the ordered nucleotides (1 to read length) representing the
abscissa, and quality values representing the ordinate. In this
research we model read quality values as polynomial functions obtained
with least-squares fitting, as one approach to compression read
quality values by modeling.

@@

$k$-means clustering is a vector quantization method, partitioning a
set of samples into $k$ sets that minimize within-cluster sum of
squares~\citep{macqueen1967some}. Using a random subset of read
quality values, a compression method can use the computed cluster
centers as read quality profiles. As the problem is NP-hard, we use a
heuristic iterative refinement approach quickly converging to a
locally optimal minimum provided by the Python package scikit-learn
0.15~\citep{pedregosa2011scikit}.

@@

In general, the greater the polynomial order, the better overall
assembly; however, the 5th-degree polynomial regression performs
slightly worse than the 3rd-degree polynomial. The respective ranks in
terms of N50 and relocations are fairly distant, which lowers the
overall ranking of the 5th-degree polynomial slightly below that for
3rd-degree polynomial model. The 1st- and 0-degree polynomial
regression methods perform poor in all metrics except assembled bases.
One explanation is that the high error portions of sequences are being
marked as high quality, so \textsc{allpaths-lg} is unable to trim or
error correct the sequences. Assembled sequences that overlap maybe
unable to align to the errors at the end of the reads, artificially
inflating the assembled genome size.

>>>>

\section{Results:}
Lossy compression of quality information can greatly decrease storage
and memory requirements with little discernible effects on subsequent
analysis results. The three compression strategies in this study were
able to produce similar results to those obtained with uncompressed
quality sequences in terms of quality control, genome assembly, and
alignment of short read to a reference sequence.

@@

If quality values are modeled, compression is conceivably possible by
replacing the original quality values by a representation of the
model. For example, quality values can be conceptualized as bivariate
data with the ordered nucleotides (1 to read length) representing the
abscissa, and quality values representing the ordinate. In this
research we model read quality values as polynomial functions obtained
with least-squares fitting, as one approach to compression read
quality values by modeling.

@@

$k$-means clustering is a vector quantization method, partitioning a
set of samples into $k$ sets that minimize within-cluster sum of
squares~\citep{macqueen1967some}. Using a random subset of read
quality values, a compression method can use the computed cluster
centers as read quality profiles. As the problem is NP-hard, we use a
heuristic iterative refinement approach by quickly converging to a
locally optimal minimum provided by R~\citep{hartigan1979algorithm}.

@@

In general, the greater the polynomial order, the better overall
assembly; however, the 5th-degree polynomial regression performs
slightly worse than the 3rd-degree polynomial. The respective ranks in
terms of N50 and relocations are fairly distant, which lowers the
overall ranking of the 5th-degree polynomial slightly below that for
3rd-degree polynomial model. The 1st- and 0-degree polynomial
regression methods perform poor in all metrics except assembled bases.
One explanation is that the high error portions of sequences are being
marked as high quality, so \textsc{allpaths-lg} is unable to trim or
error correct the sequences. Assembled sequences that overlap maybe
unable to align across the errors at the end of the reads, artificially
inflating the assembled genome size.
